Verifying Concurrent Search Structure Templates for C Programs using VST

### 2023-04-25 concurrency/CONCUR 2023/paper.tex
filling in some of my pieces

The Concurrent Search Structure Template approach~\cite{templates} aims make the problem tractable by separating verification of concurrency control patterns from verification of the underlying data structure. Addressing the proof of each component separately makes the individual proofs easier, and also offers the possibility of proof reuse: each verified concurrency pattern (``template'') can be applied to many different data structures, and each verified sequential data structure can be outfitted with many different templates. In theory, verifying $n$ templates and $m$ data structures might yield $n \times m$ verified concurrent data structures; in practice, as we will see, both templates and data structures make assumptions that may invalidate certain combinations. %how much more can we say about this?

We will discuss several templates in section~\ref{templates}, but broadly, a search structure template consists of a \lstinline{traverse} function that travels through a data structure in search of a particular key, using the appropriate concurrency control mechanism when moving from node to node. The node to travel to is selected by a black-box function \lstinline{findNext} provided by the data structure; all the template needs to know is that it has some way of choosing a next node to examine when the current node does not contain the target key. \wm{Show an example template here?} Once the appropriate node for the key has been found, the template returns it to a top-level function that calls out to the data structure to perform the appropriate operation (e.g. insert, lookup, delete) on that node. Thus, the \lstinline{traverse} function can be verified without knowing anything about the target data structure other than its synchronization mechanism, and without knowing anything about the operation to be performed other than that it can be performed on a single node (or sometimes a node and its parent). The job of the data structure is to implement the \lstinline{findNext} function and the specific operations; the combination of the template and the data structure is then a fully operational concurrent data structure. %what more do we want to say here?

\subsection{Iris and VST}
\than{write more here ... }

@@

\section{Verified Data Structures}
In this section, we demonstrate how to integrate all components into high-level specifications for the purpose of verifying two template algorithms, lock-coupling and give-up, in the real C programming language.

\than{Write more here ... }

@@

\subsection{Templates vs. Internal Reorganization} % move to section 4?
In the binary search tree, \texttt{pushdown\_left} involves both the structure of the tree and the concurrency protocol. There's no good way to disentangle this into a template and a concurrency-free data structure operation.

\than{Write more ...}

>>>>

The Concurrent Search Structure Template approach of Krishna et al.~\cite{templates} aims make the problem tractable by separating verification of concurrency control patterns from verification of the underlying data structure. Addressing the proof of each component separately makes the individual proofs easier, and also offers the possibility of proof reuse: each verified concurrency pattern (``template'') can be applied to many different data structures, and each verified sequential data structure can be outfitted with many different templates. In theory, verifying $n$ templates and $m$ data structures might yield $n \times m$ verified concurrent data structures; in practice, as we will see, both templates and data structures make assumptions that may invalidate certain combinations. %how much more can we say about this?

We will discuss several templates in section~\ref{templates}, but broadly, a search structure template consists of a \lstinline{traverse} function that travels through a data structure in search of a particular key, using the appropriate concurrency control mechanism when moving from node to node. The node to travel to is selected by a black-box function \lstinline{findNext} provided by the data structure; all the template needs to know is that it has some way of choosing a next node to examine when the current node does not contain the target key. \wm{Show an example template here?} Once the appropriate node for the key has been found, the template returns it to a top-level function that calls out to the data structure to perform the appropriate operation (e.g. insert, lookup, delete) on that node. Thus, the \lstinline{traverse} function can be verified without knowing anything about the target data structure other than its synchronization mechanism, and without knowing anything about the operation to be performed other than that it can be performed on a single node (or sometimes a node and its parent). The job of the data structure is to implement the \lstinline{findNext} function and the specific operations; the combination of the template and the data structure is then a fully operational concurrent data structure. Krishna et al presented templates in terms of \emph{flow interfaces}~\cite{krishna2017flow}, a framework for reasoning about graph-structured data structures, but many common search structures are not graphs; in this paper, we present a version of templates that is independent of flow interfaces. %is this a contribution in itself?
%what more do we want to say here?

\subsection{Iris and VST}
The search structure template approach is built on \emph{concurrent separation logic}, which is used to specify and prove pre- and postconditions for the template and data structure functions. Krishna et al. implemented their framework in Iris~\cite{iris}, a language-independent separation logic framework built in the Coq proof assistant~\cite{coq}, with flexible support for ghost state, invariants, and atomic specifications. This makes it easy to describe the effects of concurrency control functions independently of the underlying data structure, e.g., as ``this function atomically finds a node that contains the target key.'' %forward reference to where we show the specs? or show them here?
The existing implementation is written in Iris's HeapLang, a simple functional programming language with shared-memory concurrency.

To apply the template approach to real-world code, we instead use the Verified Software Toolchain (VST)~\cite{plcc}, a separation-logic-based verifier for C programs. VST is also built in Coq and is connected to the CompCert verified C compiler~\cite{compcert}, giving it a realistic semantics and a guarantee that proved properties will hold on compiled code. Recent work on VST~\cite{vst-iris} extended it to support most of the advanced concurrency features of Iris, including ghost state, invariants, and atomic specifications. Building on this, we were able to reconstruct the template approach in VST and apply it to real C programs. %clean this up a little

@@

\section{Verified Data Structures}
In this section, we demonstrate how to integrate all components into high-level specifications for the purpose of verifying two template algorithms, lock-coupling and give-up, in the real C programming language. The target specifications for our search structure(s) are shown below: each operation should atomically have its effect on the data structure, modifying it or returning information as appropriate\footnote{Krishna et al.~\cite{templates} combine these specifications into a single function parameterized by the choice of operation; however, clients will still ultimately want to call \lstinline{insert}, \lstinline{lookup}, or \lstinline{delete} functions.}.

@@

\subsection{Templates vs. Internal Reorganization}
Our original goal was to verify an existing concurrent BST implementation with the template approach, refactoring as necessary but otherwise leaving the code unchanged. However, there is a fundamental mismatch between the template approach and the standard approach to deletion in a BST. The operations we have seen thus far can be factored into a data-structure-agnostic concurrent step (\lstinline{traverse}) and a concurrency-unaware data structure step (\lstinline{insert_op} or \lstinline{lookup_op}). When removing a node from a BST, on the other hand, we usually restructure the tree with a \lstinline{pushdown} operation, %include figure?
rotating three-node sections of the tree until the node to be deleted is at a leaf. This operation is simultaneously concurrency-aware (we must lock the nodes involved to avoid race conditions) and data-structure-specific (we need to know precisely which nodes to target and how to rearrange them). In fact, most of the existing template proofs either do not include deletion at all, or do a very lightweight form of deletion in which nodes are marked as ``deleted'' but are never actually removed from the data structure.

We believe this represents a fundamental limitation in the template approach. Operations in the style of \lstinline{pushdown} are ubiquitous in tree structures (e.g., the rebalancing operations of AVL and red-black trees), and always weave together synchronization and restructuring in a way that cannot easily be decomposed. This is a serious impediment to the goal of verifying concurrency control separately from the data structure: even if we have a verified template and a verified single-threaded tree, we still need to verify a complex data-structure-specific concurrent function. %in fact, usually the most complicated part!
%what's the conclusion here? can we see a way through?
